Complete draft

### 2019-05-08 doc/point.tex
version for Pat

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{physics}
\title{Complete draft}
%\author{Chao Ju, Alex Kim}
\begin{document}
\maketitle

There is sufficient information in the distribution of peculiar velocities and overdensities of galaxies to test general relativity on the largest scale. Suppose we have a region that encloses $N_{gal}$ number of galaxies in the universe. We can divide the region into $N_{cell}\gg 1$ number of equal volume cells whose size can be made so small that either one or no galaxy is contained in one cell. This is encoded in a vector $\textbf{N}$ of size $N_{cell}$ whose entries are either 0 or 1. For any 2 cells, we can compute their correlation, a function of the distance $r$ between the cells only. In our application, we consider $\xi_{\delta\delta}$ (the overdensity-overdensity correlation), $\xi_{\delta v}$ (overdensity-velocity correlation), $\xi_{v\delta}$ (velocity-overdensity correlation), and $\xi_{vv}$ (velocity-velocity correlation).\par
We postulate that deviations from expected variance $\bar{N}$ and expected peculiar velocity $\bar{v}$ are Gaussian. Then the above correlation matrices can be used to construct a block matrix $\textbf{C}$, parameterized by $A$ and $\bm{\sigma}_v$, a vector of shot noise for peculiar velocity measurement:
\begin{align}
\textbf{C} &\equiv \textbf{K} + A\textbf{S} \\ 
\bar{N}\textbf{I} & 0 \nonumber \\
0 & \bm{\sigma}_v^2\textbf{I}
+
A
\bar{N}^2\xi_{\delta\delta} & \bar{N}\xi_{\delta v} \\
\bar{N}\xi_{\delta v} & \xi_{vv}
\right).
\end{align}\par
The reason we have $\bar{N}$ as the variance for overdensity is that our small galaxy cells are best modeled by Poisson statistics, in which the mean equals the variance. In addition, note that the dimension of $C$ is $N_{cell}+N_{gal}$ by $N_{cell}+N_{gal}$. For the peculiar velocity covariance, we assign $\sigma^2=\infty$ to cells that do not contain any galaxy, and one can show that removing rows and columns that contain those cells has no effect but a constant offset on what we will compute later. \par
Given the block matrix $C$ and our data $\textbf{x}= [N_1-\bar{N}, N_2-\bar{N}, \hdots, N_{N_{cell}}-\bar{N}, \Delta v_1, \hdots, \Delta v_{N_{gal}}]^T$, where $\Delta v$ is the deviation of the measured velocity from the expected, we can form the Gaussian likelihood as a function of $A$
\mathcal{L}(A) = \frac{1}{\sqrt{\left(2\pi\right)^{N_{cell}+N_{gal}} \det \textbf{C}}}\exp\left\{-\frac{1}{2}\textbf{x}^T\textbf{C}^{-1}\textbf{x}\right\}.
To make the expression cleaner, we take the log of the likelihood and focus on the $A$-dependent part:
-2\ln\mathcal{L}(A) = \ln\det\textbf{C} + \textbf{x}^T\textbf{C}^{-1}\textbf{x} + \text{const}.
\end{equation}\par

@@

\begin{thebibliography}{9}
\bibitem{adam}
C. Adams, C. Blake. 
Improving constraints on the growth rate of structure by modelling the density-velocity cross-correlation in the 6dF Galaxy Survey. arXiv:1706.05205 [astro-ph.CO].
\end{thebibliography}

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{physics}
\usepackage{verbatim}
\title{Complete draft}
%\author{Chao Ju, Alex Kim}
\begin{document}
\maketitle

There is sufficient information in the distribution of peculiar velocities and overdensities of galaxies to test general relativity on the largest scale. Suppose we have a region that encloses $N_{gal}$ number of galaxies in the universe. We can divide the region into $N_{cell}\gg 1$ number of equal volume cells whose size can be made so small that either one or no galaxy is contained in one cell. This is encoded in a vector $\textbf{N}$ of size $N_{cell}$ whose entries are either 0 or 1. For any 2 cells, we can compute their correlation, a function of the distance $r$ between the cells only. In our application, we consider $\xi_{\delta\delta}$ (the overdensity-overdensity correlation), $\xi_{\delta v}$ (overdensity-velocity correlation), $\xi_{v\delta}$ (velocity-overdensity correlation), and $\xi_{vv}$ (velocity-velocity correlation).\par
We postulate that deviations from expected variance $\bar{N}$ and expected peculiar velocity $\bar{v}$ are Gaussian. Then the above correlation matrices can be used to construct a block matrix $\textbf{C}$, parameterized by $A$ and $\bm{\sigma}_v$, a vector of shot noise for peculiar velocity measurement:
\begin{align}
\textbf{C} &\equiv \textbf{K} + A\textbf{S} \\ 
&=
\left(
\begin{array}{cc}
\bar{N}\textbf{I} & 0 \nonumber \\
0 & \bm{\sigma}_v^2\textbf{I}
\end{array}
\right)
+
A
\left(
\begin{array}{cc}
\bar{N}^2\xi_{\delta\delta} & \bar{N}\xi_{\delta v} \\
\bar{N}\xi_{\delta v} & \xi_{vv}
\end{array}
\right).
\end{align}\par
The reason we have $\bar{N}$ as the variance for overdensity is that our small galaxy cells are best modeled by Poisson statistics, in which the mean equals the variance. In addition, note that the dimension of $C$ is $N_{cell}+N_{gal}$ by $N_{cell}+N_{gal}$. For the peculiar velocity covariance, we assign $\sigma^2=\infty$ to cells that do not contain any galaxy, and one can show that removing rows and columns that contain those cells has no effect but a constant offset on what we will compute later. \par
Given the block matrix $C$ and our data $\textbf{x}= [N_1-\bar{N}, N_2-\bar{N}, \hdots, N_{N_{cell}}-\bar{N}, \Delta v_1, \hdots, \Delta v_{N_{gal}}]^T$, where $\Delta v$ is the deviation of the measured velocity from the expected, we can form the Gaussian likelihood as a function of $A$
\[
\mathcal{L}(A) = \frac{1}{\sqrt{\left(2\pi\right)^{N_{cell}+N_{gal}} \det \textbf{C}}}\exp\left\{-\frac{1}{2}\textbf{x}^T\textbf{C}^{-1}\textbf{x}\right\}.
\]\par
To make the expression cleaner, we take the log of the likelihood and focus on the $A$-dependent part:
\begin{equation}
-2\ln\mathcal{L}(A) = \ln\det\textbf{C} + \textbf{x}^T\textbf{C}^{-1}\textbf{x} + \text{const}.
\label{log:eqn}
\end{equation}\par

%  ---- Change
%NEW STUFF

We new expand the log-likelihood as a Taylor series for small $A$ around $A=0$.

First, we focus on the  term $\textbf{x}^T\textbf{C}^{-1}\textbf{x}$. We rewrite $\textbf{C}$ by normalizing out its uncorrelated $\textbf{K}$ term
by   introducing the matrices $\textbf{M}$ and $\textbf{X}$,
where $\textbf{M}=\sqrt{\textbf{K}}$ and $\textbf{S} = \textbf{M} \textbf{X} \textbf{M}$ such that $\textbf{X}=\textbf{M}^{-1}\textbf{S}\textbf{M}^{-1}$. Now, our matrix $\textbf{C}$ can be re-written in terms of $\textbf{X}$ and $\textbf{M}$:
\begin{equation}
\textbf{C} = \textbf{M}(\textbf{I}+A\textbf{X})\textbf{M}.
\end{equation}
Expanded as a Taylor series
\begin{align}
\textbf{C}^{-1} & = \textbf{M}^{-1}(1-A\textbf{X}+A^2 \textbf{X}^2 + \ldots) \textbf{M}^{-1} \\
& = \textbf{K}^{-1}   -A \textbf{K}^{-1}\textbf{S}\textbf{K}^{-1} +\frac{A^2}{2} \textbf{K}^{-1}\textbf{S}\textbf{K}^{-1}\textbf{S}\textbf{K}^{-1} +O(A^3) + \ldots 
\end{align}
%= -A \textbf{K}^{-1}\textbf{S}\textbf{K}^{-1} + \text{const}.
%\] \par

We evaluate $\textbf{x}^T \textbf{C}^{-1}\textbf{x}$ by splitting up the counts and velocities.  The data vector $\textbf{x}$ is composed of the counts
 $\textbf{N}-\bar{N}$ and peculiar velocities $\mathbf{\Delta v}$, whereas $\textbf{S}$ is comprised of $\bar{N}^2\xi_{\delta\delta}$, $\bar{N}\xi_{\delta v}$,
and $\xi_{vv}$.  In this manner
\begin{align}
\textbf{x}^T \textbf{K}^{-1}\textbf{S}\textbf{K}^{-1} \textbf{x}& = (\textbf{N}-\bar{N}) \xi_{\delta\delta}  (\textbf{N}-\bar{N})  + 2  (\textbf{N}-\bar{N}) \xi_{\delta v}\frac{\mathbf{\Delta v}}{\bm{\sigma}_v^2}+ 
\frac{\mathbf{\Delta v}}{\bm{\sigma}_v^2} \xi_{vv}\frac{\mathbf{\Delta v}}{\bm{\sigma}_v^2}.
\end{align}
Noting that $N_i=1$ when the cell $i$ contains a galaxy and zero otherwise and that only cells with a peculiar velocity measurement contribute ($\sigma_v =\infty$
for cells without such a measurement), and that in the limit of infinitely small cells, $\bar{N}\sum_{i,j\in N_{cell}} = n \int dV$, where $n$ is the galaxy number density,
\begin{align*}
\textbf{x}^T \textbf{K}^{-1}\textbf{S}\textbf{K}^{-1} \textbf{x}& = \sum_{i,j\in N_{gal}} \xi_{\delta\delta,ij} -2 \bar{N} \sum_{i\in N_{gal},j\in N_{cell}} \xi_{\delta\delta,ij}\\
& +\bar{N}^2  \sum_{i,j\in N_{cell}}  \xi_{\delta\delta,ij} + 2 \sum_{i,j\in N_{gal}} \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,ij} \\
& -2\bar{N}\sum_{i\in N_{cell}, j\in N_{gal}} \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,ij}  + \sum_{i,j\in N_{gal}} \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{vv,ij} \frac{\Delta v_j}{\sigma_{v,j}^2}\\
& = \sum_{i,j\in N_{gal}} \left( \xi_{\delta\delta,ij} + 2  \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,ij} + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{vv,ij} \frac{\Delta v_j}{\sigma_{v,j}^2}\right)\\
&-2 n \sum_{i\in N_{gal}}\int \left(\xi_{\delta\delta,iV} + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{v \delta,iV}\right) dV\\
& +n^2  \int \xi_{\delta\delta,V_1V_2}dV_1dV_2.
\end{align*}

Now consider the next term $\textbf{K}^{-1}\textbf{S}\textbf{K}^{-1}\textbf{S}\textbf{K}^{-1}$.  The middle section is
\begin{align}
\textbf{S}\textbf{K}^{-1}\textbf{S} & =
\begin{pmatrix}
 \bar{N}^2  \xi_{\delta v}\frac{1}{\bm{\sigma}_v^2}  \xi_{v \delta}  & \bar{N} \xi_{\delta v} \frac{1} {\bm{\sigma}_v^2} \xi_{vv}  \\
 \bar{N} \xi_{v v} \frac{1} {\bm{\sigma}_v^2} \xi_{v \delta}   &  \xi_{vv} \frac{1}{\bm{\sigma}_v^2} \xi_{vv}
\end{pmatrix}
\end{align}
to leading order of $\bar{N}$ anticipating that we will take the limit where it goes to zero.
Then
\begin{align}
\textbf{x}^T \textbf{K}^{-1}\textbf{S}\textbf{K}^{-1}\textbf{S}\textbf{K}^{-1} \textbf{x}
& = \sum_{i,j\in N_{gal}} \left( \left(\xi_{\delta v}\frac{1}{\bm{\sigma}_v^2}  \xi_{v \delta}\right)_{ij}
 + 2  \frac{\Delta v_j}{\sigma_{v,j}^2} \left(\xi_{\delta v} \frac{1} {\bm{\sigma}_v^2} \xi_{vv}  \right)_{ij} + \frac{\Delta v_i}{\sigma_{v,i}^2} \left( \xi_{vv} \frac{1}{\bm{\sigma}_v^2} \xi_{vv}\right)_{ij} \frac{\Delta v_j}{\sigma_{v,j}^2}\right)\\
&-2 n \sum_{i\in N_{gal}}\int \left( \left(\xi_{\delta v}\frac{1}{\bm{\sigma}_v^2}  \xi_{v \delta}\right)_{iV} + \frac{\Delta v_i}{\sigma_{v,i}^2} \left( \xi_{v v} \frac{1} {\bm{\sigma}_v^2} \xi_{v \delta}  \right)_{iV} \right) dV\\
& +n^2  \int  \left(\xi_{\delta v}\frac{1}{\bm{\sigma}_v^2}  \xi_{v \delta}\right)_{V_1V_2}dV_1dV_2.
\end{align}

We have assembled the contributions of $\textbf{x}^T \textbf{C}^{-1}\textbf{x}$ to second order in $A$.

Returning to the other contribution to the log-likelihood in Eq.~\ref{log:eqn} we have
\begin{align*}
\ln\det(\textbf{C}) &=\ln\det(\textbf{M}(\textbf{I}+A\textbf{X})\textbf{M})) \\
&= \ln\det(\textbf{I}+A\textbf{X}) +\text{const} \\
&= A \tr \textbf{X} -\frac{A^2}{2}\tr \textbf{X}^2+ O(A^3) + \ldots\\
\end{align*}
Now
\begin{equation}
\textbf{X} = \textbf{M}^{-1}\textbf{S}\textbf{M}^{-1} = 
\begin{pmatrix}
\bar{N}\xi_{\delta\delta} & \frac{\sqrt{\bar{N}}}{\sigma_{v}} \xi_{\delta v} \\
\frac{\sqrt{\bar{N}}}{\sigma_{v}} \xi_{v\delta} & \frac{1}{\sigma_{v}^2} \xi_{vv} 
\end{pmatrix}.
\end{equation}
The contribution to the trace of the upper diagonal in the limit of small cell volumes is
\begin{equation}
\bar{N} \tr \xi_{\delta\delta} = n \int \xi_{\delta\delta,VV} dV
\end{equation}
and the lower diagonal is
\begin{equation}
\tr  \frac{1}{\sigma_{v}^2} \xi_{vv} =  \sum_{i \in N_{gal}}   \frac{1}{\sigma_{v,i}^2} \xi_{vv,ii}.
\end{equation}

The trace of $\textbf{X}^2$ has terms, to leading order in $\bar{N}$, has terms
\begin{equation}
\bar{N}  \tr  \xi_{\delta v} \frac{1}{\sigma_{v}}  \frac{1}{\sigma_{v}}  \xi_{v \delta} = n \int \left( \xi_{\delta v} \frac{1}{\sigma_{v}}  \frac{1}{\sigma_{v}}  \xi_{v \delta} \right)_{VV} dV
\end{equation}
and 
\begin{equation}
\tr   \xi_{vv} \frac{1}{\sigma_{v}^2}  \frac{1}{\sigma_{v}^2} \xi_{vv} =  \sum_{i \in N_{gal}} \left(  \xi_{vv} \frac{1}{\sigma_{v}^2}  \frac{1}{\sigma_{v}^2} \xi_{vv} \right)_{ii}
\end{equation}

Putting everything together, the expansion of the log-likelihood in $A$ has the first order coefficient
\begin{align}
c_1 &=- \sum_{i,j\in N_{gal}} \left( \xi_{\delta\delta,ij} + 2  \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,ij} + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{vv,ij} \frac{\Delta v_j}{\sigma_{v,j}^2}\right)\\
& +\sum_{i \in N_{gal}} \left( \frac{1}{\sigma_{v,i}^2} \xi_{vv,ii}   + 2n \int \left(\xi_{\delta\delta,iV} + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{v \delta,iV}\right) dV  \right)\\
& +n \int \xi_{\delta\delta,VV} dV  -n^2  \int \xi_{\delta\delta,V_1V_2}dV_1dV_2
\end{align}
and the second order coefficient
\begin{align}
c_2 & = \frac{1}{2} \Biggl( 
 \sum_{i,j\in N_{gal}} \left( \left(\xi_{\delta v}\frac{1}{\bm{\sigma}_v^2}  \xi_{v \delta}\right)_{ij}
 + 2  \frac{\Delta v_j}{\sigma_{v,j}^2} \left(\xi_{\delta v} \frac{1} {\bm{\sigma}_v^2} \xi_{vv}  \right)_{ij} + \frac{\Delta v_i}{\sigma_{v,i}^2} \left( \xi_{vv} \frac{1}{\bm{\sigma}_v^2} \xi_{vv}\right)_{ij} \frac{\Delta v_j}{\sigma_{v,j}^2}\right)\\
&- \sum_{i \in N_{gal}}\left( \left(  \xi_{vv} \frac{1}{\sigma_{v}^2}  \frac{1}{\sigma_{v}^2} \xi_{vv} \right)_{ii}  +2n \int \left( \left(\xi_{\delta v}\frac{1}{\bm{\sigma}_v^2}  \xi_{v \delta}\right)_{iV} + \frac{\Delta v_i}{\sigma_{v,i}^2} \left( \xi_{v v} \frac{1} {\bm{\sigma}_v^2} \xi_{v \delta}  \right)_{iV} \right) dV  \right)\\
&-n  \int \left( \xi_{\delta v} \frac{1}{\sigma_{v}}  \frac{1}{\sigma_{v}}  \xi_{v \delta} \right)_{VV} dV  +n^2  \int  \left(\xi_{\delta v}\frac{1}{\bm{\sigma}_v^2}  \xi_{v \delta}\right)_{V_1V_2}dV_1dV_2 \Biggr).
\end{align}

In this second-order approximation of the log-likelihood, the minimum occurs at $A_{\text{min}}=-c_1/2c_2$ and the second derivative is $2c_2$.

\begin{comment}
END NEW STUFF

@@

\end{comment}
\begin{thebibliography}{9}
\bibitem{adam}
C. Adams, C. Blake. 
Improving constraints on the growth rate of structure by modelling the density-velocity cross-correlation in the 6dF Galaxy Survey. arXiv:1706.05205 [astro-ph.CO].
\end{thebibliography}
